# Engineering Statistics
IMSE 4410 Spring 2016. Copyright 2013-2016 by Timothy Middelkoop License CC by SA 3.0

## Hypothesis Testing Part II - Chapter 9

HO: "You can't prove anything, you can only disprove it"

Procedure (Section 9-1.6):

1. Identify parameter of interest.
2. Formulate H0
3. Formulate H1
4. Determine test statistic.
5. State H0 rejection criteria.
6. Calculate
7. Reject H0 or fail to reject H0.
8. Conclusions

### Tests on the mean of a normal distribution, variance known.

Steps 1-5

* Tests on the mean of a normal distribution, variance known.
* Test: H0: $\mu=\mu_0$; H1: $\mu\ne\mu_0$
* Distribution: Xbar is the standard normal with standard deviation sigma/sqrt(n)
* Test Statistic: $Z={\bar X - \mu_0 \over \sigma/\sqrt{n} }$
* Two sided p-value:
```{r eval=FALSE}
pv <- 2*pnorm((xb-mu0)/(sigma/sqrt(n)),lower.tail=FALSE) ; pv
```
* Reject H0: p-value > $\alpha$.

```{r}
# Example 9.2 (6e): Propellant burning calculations (Step 6)
alpha <- 0.05
sigma <- 2
xb <- 51.3
n <- 25
mu0 <- 50
pv <- 2*pnorm((xb-mu0)/(sigma/sqrt(n)),lower.tail=FALSE) ; pv ## [1] 0.00115405

# p < 0.05 so we reject H0 (Step 7)
# The mean burning rate is different from 50 cm/s (Step 8)
```


What is the significance of the test? $\alpha=0.05$

* What is P(Type II error)?
* Accept H0, H0 is false
* If you have to figure this out, use the non-standard normal functions. Otherwise just use the formula.

```{r}
## Example 9-3 (6e): Type II errors.
alpha <- 0.05
mu0 <- 50.0
mu1 <- 49.0
sigma <- 2
n <- 25

## Type II error for above hypothesis test.
beta <- pnorm(qnorm(alpha/2,lower.tail=FALSE)-((mu0-mu1)*sqrt(n))/sigma) -
  	pnorm(qnorm(alpha/2,lower.tail=TRUE)-((mu0-mu1)*sqrt(n))/sigma) ; beta
# [1] 0.294582

## This is an example were you should think about where this came from!
## Be prepared to demonstrate this on an exam.

## Find the acceptance region using the non-standard normal for H0:mu0
crl <- qnorm(alpha/2,mean=mu0,sd=sigma/sqrt(n),lower.tail=TRUE) ; crl
cru <- qnorm(alpha/2,mean=mu0,sd=sigma/sqrt(n),lower.tail=FALSE) ; cru
# [1] 49.21601
# [1] 50.78399

## Now apply this to the shifted H1:mu1
beta <- pnorm(cru,mean=mu1,sd=sigma/sqrt(n),lower.tail=TRUE) - pnorm(crl,mean=mu1,sd=sigma/sqrt(n),lower.tail=TRUE) ; beta ## [1] 0.294582

## Did that make more sense?

## For good measure lets do this for the p-values as well.
2*pnorm(xb,mean=mu0,sd=sigma/sqrt(n),lower.tail=FALSE)

## Sample size given beta.

## In the case we would like a power of 0.90 (beta=0.10), what would be n
## From the text
beta <- (1-0.90)
n <- (qnorm(alpha/2,lower.tail=FALSE) + qnorm(beta,lower.tail=FALSE))^2*sigma^2/((mu0-mu1))^2 ; n

## For the computational people.
## Just numerically solve for n on the interval c(1,100)
beta <- (1-0.90)
## uniroot finds a root (a zero) of a function
uniroot(function(n) {
		-beta +
        pnorm(qnorm(alpha/2,lower.tail=FALSE)-((mu0-mu1)*sqrt(n))/sigma) -
		pnorm(qnorm(alpha/2,lower.tail=TRUE)-((mu0-mu1)*sqrt(n))/sigma)
},c(1,100))$root

## Using OC Curves. Don't. Use a computer.
```


### Tests on the mean of a normal distribution, variance unknown (9-3)

* Tests on the mean of a normal distribution, variance unknown.
* Test: H0:$\mu=\mu_0$; H1:$mu\ne\mu_0$
* Distribution: $t$ distribution with $n-1$ degrees of freedom.
* Test statistic: $T0= { \bar X - \mu_0 \over S/\sqrt{n} }$
* Two sided p-value: (we can do this since the t-dist is symmetric)
```{r eval=FALSE}
pv <- 2*pt(abs(xb-mu0)/(s/sqrt(n)),n-1,lower.tail=FALSE) ; pv 
```
* One sided p-vale, H1:$\mu>\mu_0$, upper tailed test, critical region on right
```{r eval=FALSE}
pv <- pt((xb-mu0)/(s/sqrt(n)),n-1,lower.tail=FALSE) ; pv
```
* One sided p-vale, H1:$\mu<\mu_0$, lower tailed test, critical region on left
```{r eval=FALSE}
pv <- pt((xb-mu0)/(s/sqrt(n)),n-1,lower.tail=TRUE) ; pv
```

Example:
```{r}
## Example 9-6: Golf club
ch09 <- read.csv("data/5e/ch09.csv",header=TRUE)
d <- na.omit(ch09$Example9.6) ; n <- length(d) ; n ## [1] 15
xb <- mean(d); xb ## [1] 0.83724
s <- sd(d); s ## [1] 0.0245571

## Data matches the text.
qqnorm(d,datax=TRUE); qqline(d,datax=TRUE)

## Does the mean exceed 0.82 with a significance level of 0.05 
## H0: mu==0.82; H1: mu > 0.82; upper tailed test
alpha <- 0.05
mu0 <- 0.82
pv <- pt((xb-mu0)/(s/sqrt(n)),n-1,lower.tail=FALSE) ; pv ## [1] 0.008313369

## FYI/ the t0-value
(xb-mu0)/(s/sqrt(n)) ## [1] 2.718979

## p < 0.05 so we reject H0, mu > 0.82

## R direct analysis
t.test(d, mu=0.82, alternative="greater")
```


Type II error.
Messy, so we use R.

```{r}
## Example 9-7 (6e)
n <- 15 ; alpha <- 0.05
power <- 0.80 ; delta <- 0.02
## How many samples would be needed to have a power of 0.80 for a delta of 0.02
power.t.test(n=NULL,power=power,sd=s,delta=0.02,sig.level=alpha, type="one.sample", alternative="one.sided")

## Only 10, so 15 is sufficient.

## What is the power of a test a n=15?
power.t.test(n=n,power=NULL,sd=s,delta=0.02,sig.level=alpha, type="one.sample", alternative="one.sided")

## The power is 0.91
```

The "one.sample" indicates that we are taking a single sample from
the population and trying to determine if it is from that population.
It is the name of this hypothesis test.

