# Engineering Statistics
IMSE 4410 Spring 2016. 
Copyright 2013-2016 by Timothy Middelkoop License CC by SA 3.0

## Review of Continuous Random Variables - Chapter 4

### Uniform Random Variables.

```{r}
# What is a "random" number.  Dice is discrete, uniform. Here is the continuous version.
# 100 (r)andom (unif)orm numbers stored in x
x <- runif(20) ; x
# visualize
plot(x)
hist(x)
# Close, lets make the numbers a bit larger 
x <- runif(100000)
h <- hist(x)
summary(x)
# Looks good

# This histogram is the discrete probably density function.
# Lets roll a continuous 6 sided dice
x <- runif(100,1,6)
hist(x,breaks=0:7)
# breaks are done explicitly
0:7
# Is is that what you expected?
x <- runif(100,1,7)
hist(x)
# probability for discrete version is 
1/6
# (d)ensity for the (unif)orm version (d=P(X==x) [density])
dunif(1:6,1,7)
# verify the property that the area under the curve=1
(1/6)*(7-1)
# ok so we did not need a calculator but you get the point.
# What is the probability of rolling a 1 or less?
# What is the probability of rolling a 6 or less?
punif(1,1,7)
punif(7,1,7)
# How about between 2 and 3

# Sound familiar?  What about the reverse, at what is P[x<=q]=0.50?
qunif(0.50,1,7)
# Look at the chart, does this make sense?

```


Review of R functions:

* runif(n,...) - random values drawn from the distribution.
* dunif(x,...) - the fx (probably density function - pdf) values
* punif(q,...) - the Fx (cumulative distribution function - cdf) values P(X<=x)
* qunif(p,...) - the inverse cdf values (quantile), find q given P(X<=q)==p

### Random Normal Distribution

$$ f(x) = {1 \over \sqrt{2\pi\sigma}} e^{-(x-\mu)^2 \over 2\sigma^2} , -\infty < x < \infty $$
parameters of $\mu$ and $\sigma$ or $N(\mu,\sigma)$

$$ E(X)=\mu, V(X)=\sigma^2$$

* Standard norm, so we can use the tables, is $\mu=0$, $\sigma^2=1$
* Random variable for the standard norm is Z, cumulative function Φ(z)=P(Z<=z)

```{r}
# The normal distribution is the rnorm series
hist(rnorm(1e6))

# What is P(Z<=1.5)
# Φ(1.5)
# Tables... 0.933139
# or use
pnorm(1.5)

# Z=(X-mu)/sigma
mu <- 5 ; sigma <- 3
# find P(X<=5.5) should be close to zero and positive.
pnorm((5.5-mu)/sigma)
# [1] 0.5661838
# Note the use of the solution in the code.  This is called a "Known Good Solution"
# No need to convert to standard norm
pnorm(5.5,mean=mu,sd=sigma)
```

### Exponential Distribution


The other real useful distribution is the exponential distribution.

* $\mu=1/\lambda$, $var=1/\lambda^2$
* Memoryless
* Units must be the same.
* $1/\lambda$ is the mean interarrival time (time between events)
* $\lambda$ is the mean arrival rate (events/time)


$$ f(x) = \lambda e^{-\lambda x} , 0 \le x < \infty $$

$\mu = E(X) = {1 \over \lambda}$ and $\sigma^2 = V(X) = {1 \over \lambda^2}$

It is important to use consistant units for $X$ and $\lambda$.

```{r}
# Example: 5 students per hour (1 every 12 min, 1 every .2 hour)
x <- rexp(1000,5)
hist(x)
# Look at the results
mean(x)
var(x)
# Now from the tables.
# What is the probablility that a student will come in in the next 6 min 
# This is .1 hours.
# P(X<=0.1)
pexp(.1,5)
pexp(.2,5)
## Does this make sence?
n <- length(x)
h <- hist(x)
names(h)
cumsum(h$counts/n)
```
