# Assignment A1 Solutions 
Assignment A1 solutions for Engineering Statistics (IMSE 4410) Spring 2016.
Copyright 2014-2016 by Timothy Middelkoop License CC by SA 3.0

Points are for reference only and indicate the points that were assigned as an exam.  This assignment is graded as Pass (10), Fail (5), and No Attempt (0)

## Statistics (11 points)

1. What is the *population* standard deviation of $9, 4, 2, 5, 10$? (2 Points)

$$\bar X={9+4+2+5+10\over5}=6$$
$$S^2={\sum (x_i-\bar X)\over n}={(9-6)^2+(4-6)^2+(2-6)^2+(5-6)^2)+(10-6)^2 \over 5}={46\over5}=9.2$$
$$S=\sqrt{9.2}=3.03$$

```{r}
(9+4+2+5+10)/5
(9-6)^2 + (4-6)^2 + (2-6)^2 + (5-6)^2 + (10-6)^2
sqrt(46/5) 

## using R
d <- c(9,4,2,5,10)
xb <- mean(d)
sqrt(sum((d-xb)^2)/length(d))

## Note the difference with the sample sd.
sd(d)
```

2.  From the tables find the following:

a. for the standard normal what is the probability that $Z \ge 0.5$? (2 points)

$$P(Z \ge 0.5)=1-P(Z \le 0.5)=1-0.6914625=0.3085375$$

```{r}
pnorm(0.5,lower.tail=FALSE)
```

b.  for the standard normal distribution, what $P(-3 \le Z \le 2)$? (2 points)

$$P(-3 \le Z \le 2)=P(Z \le 2)-(1-P(Z \le 3))=0.9772499-(1-0.9986501)=0.9759$$

```{r}
pnorm(2,lower.tail=TRUE) - pnorm(-3,lower.tail=TRUE)
```

c.  what is the $Z$ value for the lower $50\%$ tail for the standard normal distribution. (1 point)

$$P(Z)=0.5, Z=0$$

```{r}
qnorm(0.5,lower.tail=TRUE)
```


d.  what is the $Z$ value for $\alpha=0.10$ for a two sided hypothesis tests? (1 point)

$$Z_{\alpha/2}=\phi(1-{\alpha/2})=\phi(0.95)=1.65$$

```{r}
qnorm(0.10/2,lower.tail=FALSE)
```

e.  For a normally distributed random variable $X$ with $\mu= 16$ and $\sigma = 9$ what is the probability that $X \le 12$? (2 points)

$$P(X\le12)=P(Z\le{12-16\over9})=P(Z \le0.45)=1-P(Z\le0.45)=1-0.6736=0.3265$$

```{r}
pnorm(12,mean=16, sd=9, lower.tail=TRUE)
```


f.  What are the parameters of the standard normal distribution. (1 point)

*Normal distributions are defined by two parameters, the mean $(\mu)$ and the standard deviation $(\sigma)$.*

## R (7 points)

3.  How do you assign the number $3$ to the variable a. (1 point).


```{r}
a<-3
a=3
```

4.  How would you produce $100$ normally distributed samples with a mean of $4$ and standard deviation of $10$? (1 point)

```{r eval=FALSE}
rnorm(100,4,10)
```

5.  Write the R function and the proper arguments to find the following:

a.	For the standard normal distribution, what is the probability that $Z\le 1$. (1 point)

```{r}
pnorm(1)
pnorm(1,lower.tail=TRUE)
```

b.  For the normal with $\mu=3$ and $\sigma=12$ find $P(4\le Z \le 10)$.  (1 point)

```{r}
pnorm(10,3,12,lower.tail=TRUE)-pnorm(4,3,12,lower.tail=TRUE)
```

c.  The z value for the upper $0.15\%$ tail for the standard normal distribution. (2 points)

```{r}
qnorm(0.15,lower.tail=FALSE)
```

d. What is t-value for the upper tail of the t distribution for an alpha of $0.05$ with $4$ degrees of freedom (find $t_{0.05,4}$). (1 point)

```{r}
qt(0.95,4)
qt(0.05,4,lower.tail=FALSE)
```

## Confidence Intervals (6 points)

6.  What distribution, and it's parameters, is used to determine the confidence interval on the variance with unknown variance from a normal population? (4 points)

*Student* or *t*

7.  What is the margin of error for the two sided confidence interval on the mean, variance known, small sample size (in either mathematical notation or in R notation). (2 point)

$$\pm t_{\alpha/ 2} \frac{S}{\sqrt{n}}$$

## Computation (26 points)

8. For a two sided confidence interval for the mean of a population, standard deviation unknown, large sample size, the random variable, $(\bar X-\mu)\over{{S}/{\sqrt{n}}}$ , is approximately normal. The confidence interval can be found as follows: 
$$\bar X-Z_{\alpha/2}\frac{S}{\sqrt{n}}\le\mu\le\bar X+Z_{\alpha/2}\frac{S}{\sqrt{n}}$$ 
with $Z_{\alpha/2}$ being the upper tail of the standard normal distribution giving a $100(1-\alpha)\%$ confidence level.

a. What R function (with proper argument values) would you use to find $Z_{\alpha/2}$?

qnorm($\alpha/2$,lower.tail=FALSE)

b. A large sample of $32$ is taken with a sample mean of $12$ and a sample standard deviation of $2$ from a normal population with a mean of $12.5$ and standard deviation of $1.2$, compute the $90\%$ two sided confidence interval?

```{r}
xb<-12
s<-2
n<-32
me<-qnorm(0.95)*s/sqrt(n)
left<-xb-me
right<-xb+me
left
right
```

```{r}
qnorm(0.05,mean=12,sd=2/sqrt(32),lower.tail=TRUE)
qnorm(0.05,mean=12,sd=2/sqrt(32),lower.tail=FALSE)
```

9.  A sample of $c(3,2,5,10)$ was taken from a population that is normally distributed with a known standard deviation of $1.2$.  Compute the $90\%$ confidence interval on the mean. (18 Points)

```{r}
mean(c(3,2,5,10))
xb<-5
s<-1.2
n<-4
error<-qnorm(0.95)*s/sqrt(n)
left<-xb-error
right<-xb+error
left
right
```




