# Engineering Statistics
IMSE 4410 Spring 2016.  Copyright 2013-2016 by Timothy Middelkoop License CC by SA 3.0

## Design of Experiments with Several Factors - Chapter 14

```{r}
ch14 <- read.csv("data/5e/ch14.csv",header=TRUE)
nrow(ch14)
```

### Two Factorial Experiments. (14-3)

Done on the board first fully by hand.  Similar to block but extended to factor a, factor b, and replication.

Model: $$y_{ijk} = \mu + A_i + B_j + AB_{ij} + e_{ijk}$$ (different notation to the text)

SS Identity: $$SS_T=SS_A+SS_B+SS_{AB}+SS_E$$ $SS_{AB}$ is the interaction between a and b

A full factorial experiment has n replications.  Experiment *must* be done in random order.

$H_0$: no factor/effect effect, all $A_i$, $B_j$, $AB_{ij}$ is zero (see text).

$H_1$: one of them is non-zero so that factor/interaction has 
an effect.

```{r}
## Example 14-1, primer paints, method:{dipping, spraying} and primer:{1,2,3}

d <- na.omit(data.frame(y=ch14$Example14.1Adhesion,a=factor(ch14$Example14.1Primer),b=factor(ch14$Example14.1Method)))

## Use R directly
m <- aov(y~a*b,d) ; summary(m)

## Lets do this by hand.
a <- length(levels(d$a)) ; a # [1] 3
b <- length(levels(d$b)) ; b # [1] 3

## What ?!
levels(d$b)
```

What is this?  Data quality is always an issue! Check. Check. Check.

Lets try a new approach, the na.omit does not work here. Sometimes simple is not better than correct.

```{r}
## This is the correct way to block the data off, lets rename the columns at the same time.
p <- ch14[c('Example14.1Adhesion','Example14.1Primer','Example14.1Method')]
names(p) <- c('adhesion','primer','method')
N <- nrow(na.omit(p)) ; p <- p[1:N,] ; N # [1] 18

## This looks a lot cleaner now.
d <- data.frame(y=p$adhesion,a=factor(p$primer),b=factor(p$method))
m <- aov(y~a*b,d) ; summary(m)

## Interestingly enough the aov function does the right thing when handling the factor earlier.

a <- length(levels(d$a)) ; a # [1] 3
b <- length(levels(d$b)) ; b # [1] 2

## Replications per combination of factors (not N)
n <- length(d$y)/a/b ; n 
# [1] 3

yb <- mean(d$y); yb # [1] 4.988889

sst <- sum((d$y-yb)^2) ; sst # [1] 10.71778

## check with text (row)
aggregate(y~a,d,c)

## compute row averages.
yba <- aggregate(y~a,d,mean)$y ; yba

## should be the column averages
ybb <- aggregate(y~b,d,mean)$y ; ybb

## check with the text (example reports column sums)
aggregate(y~b,d,sum)$y

## ssa
ssa <- b*n*sum((yba-yb)^2) ; ssa

## ssb
ssb <- a*n*sum((ybb-yb)^2) ; ssb

## ssab  
## n * sum_a,sum_b (ybij-yba-ybj-yb)^2

## This requires the sample mean (k)
aggregate(y~a+b,d,c)

bab <- aggregate(y~a+b,d,mean) ; bab

## Don't you like it when things work the way you would expect them.
```

The problem with the next analysis is that we have a double sum with varying indexes. Again we are getting bit by "simpler is not always better than correct."  As we do more sophisticated things and start removing assumptions.

Before we were assuming that the factors where in order (they are), and are values (an i and j).  This time we need to use factors and the order of the data can be arbitrary.

```{r}
## Group the y's by common a+b.  This works since bab is produced the same
ab <- aggregate(y~a+b,d,c) ; ab
sse <- sum((ab$y - bab$y)^2) ; sse # [1] 0.9866667
```

We could continue to use this method but it becomes more difficult. See Advanced R below for details.

We can also carry the summary stats in the main data frame.
Like the strategy for sse as before with factors (not indexes) and multiple factors.

```{r}
## We use dd since we are altering our data frame.
## The hard part is to pull in the multiple columns, here we need to merge the data
dd <- d
dd$yba <- yba[d$a]
dd$ybb <- ybb[d$b]
dd$yb <- yb
dd <- merge(dd,bab,by=c('a','b'),suffixes=c('','bn')) ; dd

## Now we can find ssab (note the dropping on n multiplyer)
ssab <- sum((dd$ybn-dd$yba-dd$ybb+dd$yb)^2) ; ssab # [1] 0.2411111

## Verify solution
sst
ssa+ssb+ssab+sse # [1] 10.71778

## We could have also used this on sbb and sba (note removeal of factor before)
ssa <- sum((dd$yba-yb)^2) ; ssa # [1] 4.581111
ssb <- sum((dd$ybb-yb)^2) ; ssb # [1] 4.908889
sse <- sum((dd$y-dd$ybn)^2) ; sse # [1] 0.9866667
ssa+ssb+ssab+sse # [1] 10.71778

summary(m)
```

### Advanced R.
This is where it gets tricky.  Do not expect this to be on the exam under R. However this will give you a good idea how indexing works in R.

```{r}
## To help us along lets first do the following:
## Lets regenerate ybb again to remind us of what is going on
bb <- aggregate(y~b,d,mean) ; bb

## just show the y column
ybb <- bb$y ; ybb

## Now show the levels (Dip/Spray) for the data
d$b

## For example the we can store the 'spray' level in spray
spray <- bb[2,'b'] ; spray

## Extrat a single element
ybb[spray]

## now report the ybb for each row in b, they should match, (similar to previous method)
ybb[d$b]

## Symmetric on bb
bb$y[d$b]

## Same with the primer type.
ba <- aggregate(y~a,d,mean) ; ba

## Primer types for each row in d
d$a

## What is the corrisponding yba (similar to previous approach)
yba[d$a]
ba$y[d$a]

## Back to what we were trying to do.
## n * sum_i^a,sum_j^b (ybij-ybi-ybj-yb)^2
bab

## oh boy.
bab$y

## What is the yba for each row in bab
bab$a

yba

## Same for yba
yba[bab$a]

bab$b

ybb

ybb[bab$b]

## Here we go.
ssab <- n*sum((bab$y-yba[bab$a]-ybb[bab$b]+yb)^2) ; ssab # [1] 0.2411111
```

### 2k Factorial Design (Section 14-5)

k=3 or more

```{r}
## Signs
expand.grid(A=c('-','+'),B=c('-','+'),C=c('-','+'))

## More useful if it is +1 and -1
t <- expand.grid(A=c(-1,+1),B=c(-1,+1),C=c(-1,+1))
t$AB <- t$A*t$B
t$BC <- t$B*t$C
t$ABC <- t$A*t$B*t$C
t$I <- 1
t

## Same as the text, can you determine the Treatment Combination?

## Note the properties, equal + and - except the identity, columns are orthogonal.
## A^2 B^2 C == C since any column multiplied by itself is I
t$AB*t$ABC

## Load data, Example 14.2
d <- ch14[c('Example14.2Roughness','Example14.2Feed','Example14.2Depth','Example14.2Angle')]
names(d) <- c('y','A','B','C')
N <- nrow(na.omit(d)) ; d <- d[1:N,] ; N # [1] 16
d

## Again, simple problem, interesting data manipulation.
s <- aggregate(y~A+B+C,d,sum) ; s 

## check it matches the table in the text, same order, no need to change.

## Main effect of A is Contrast_A/(n2^(k-1))
k <- 3; n <- 2 ; n*2^(k-1)

sum(s$y*t$A)/(n*2^(k-1)) # [1] 3.375

## Sum Squares (Contrast_A)^2/(n2^k)
n*2^k # [1] 16

sum(s$y*t$A)^2/(n*2^k) # [1] 45.5625

## This is a simple Regression/ANOVA
m <- aov(y~A*B*C,d) ; summary(m)
```

We are not going to bother with regression as we can find it
with the contrasts and R does not easily compute it directly.
F-values are similar in nature to the t-stats.

```{r}
## Residual analysis with A, B and AB
m <- aov(y~A+B+A*B,d) ; summary(m)
plot(m)

## Note we can get the only the primary and secondary effects if we do the following
summary(aov(y~(A+B+C)^2,d))
```
See the manual for more details on formulas.

```{r}
## Example 14.4
d1 <- c(9,10,9,12,11,10,10,16)
d2 <- c(7,12,11,15,10,13,8,14)

d1+d2

sum(d1+d2) # [1] 177

y <- c(d1,d2) ; y 

mean(y) # [1] 11.0625

sum((y-mean(y))^2) # [1] 92.9375
```

