# Engineering Statistics
IMSE 4410 Spring 2016.  Copyright 2013-2016 by Timothy Middelkoop License CC by SA 3.0

## Analysis of Variance (AoV) - Chapter 13

```{r}
ch13 <- read.csv("data/5e/ch13.csv",header=TRUE)
nrow(ch13)
```

### Randomized Complete Block (Section 13-4)

Now the experiments are in "blocks." Additional variation is between between blocks (think different fields).  

* Each 'block' has 'a' observations indexed by $i$ and there are 'b' blocks indexed by $j$.

The *linear statistical model* is
$$Y_{ij}=\mu + \tau_i + \beta_j + \epsilon_{i,j}$$

Where $\tau_i$ is the treatment effect and $\mu$ is the grand mean and defined such that $\sum \tau_i = 0$ and $\sum \mu_j = 0$ and $\epsilon_{ij}$ is normally and independently distributed with mean of zero and variance of $\sigma^2$.

The *sum of squares identity* for the *randomized complete block design*

$$SS_T=SS_F + SS_B + SS_E$$
$$\sum_{i=1}^a\sum_{j=1}^b(y_{ij}-\bar y_{\cdot\cdot})^2 = b \sum_{i=1}^a(\bar y_{i\cdot}-\bar y_{\cdot\cdot})^2 + a \sum_{j=1}^b(\bar y_{\cdot j}-\bar y_{\cdot\cdot})^2 + \sum_{i=1}^a\sum_{j=1}^b(y_{ij}-\bar y_{i\cdot} -\bar y_{\cdot j} + \bar y_{\cdot\cdot})^2$$

*Degrees of Freedom*
$$(ab-1)=(a-1)+(b-1)+(a-1)(b-1)$$

Using this we can calculate $F_0$ and the mean squares similarly using the above information.

```{r}
d <- na.omit(data.frame(y=ch13$Example13.5Strength,a=factor(ch13$Example13.5Chemical)))
N <- nrow(d) ; N

## we now have b blocks and a treatments.
a <- length(levels(d$a)); a # [1] 4

## Look at the data in table format.
aggregate(y~a,d,c)

boxplot(y~a,d)

## Complete block so we can do
b <- N/a ; b # [1] 5

## We assume blocks are reported sequentially, add the information

## A little magic-foo.
rep(1:b,times=a)

## and some more
rep(1:a,each=b)

## Add it to our data.frame
d$b <- factor(rep(1:b,times=a)) ; d[c(1,2,19,20),]

## Now we need to extract some information, mainly ybi, ybj
## Verify with example.

## row sums like last time, sum of the treatments (x is now a).
aggregate(y~a,d,sum)

## ybi is the row average like last time, the sum of the treatments (ybi contains the y column.
ybi <- aggregate(y~a,d,mean)$y ; ybi

## now columns are just symmetric.
aggregate(y~b,d,sum)

## ybj
ybj <- aggregate(y~b,d,mean)$y ; ybj

## yb
yb <- mean(d$y) ; yb # [1] 1.96

## Should match
mean(ybj) ; mean(ybi)

## Now lets compute using the function versions
m <- aov(y~a+b,d) ; summary(m)
```

### Randomized Block Computations.

Compute by hand. Equation 13-27, sum of squares identity for randomized block. Note the split between the treatments and blocks $ssf = ssa + ssb + sse$.  We also use ssa and ssf interchangeably.

```{r}
sst <- sum((d$y-yb)^2) ; sst # [1] 25.688

## run ybi-yb to make sure it does what you expect.
ssa <- b*sum((ybi-yb)^2) ; ssa # [1] 18.044

ssb <- a*sum((ybj-yb)^2) ; ssb # [1] 6.693

## We cheat again.
sse <- sst - ssb - ssa ; sse # [1] 0.951

## How do we do this it is a bit different than before, we need to use the indexes directly for ybi and ybj. This is done by storing i and j in the data.frame.

## yij is treatment i, block j.

## j is the block
d$j <- rep(1:b,times=a) ; d$j

d$i <- rep(1:a,each=b) ; d$i

## Check
d[c(1,2,19,20),]

## Why are i and j the same as a and b ?

## Now apply it element by element.
## The term +yb comes from the fact that the "common" is removed twice/overlap.
sse <- sum((d$y-ybj[d$j]-ybi[d$i]+yb)^2) ; sse # [1] 0.951

## we can compute mse, msb, msf fromthe text to compute F0.
## Use the "new" definitions.

msf <- ssa/(a-1) ; msf # [1] 6.014667

msb <- ssb/(b-1) ; msb # [1] 1.67325

mse <- sse/((a-1)*(b-1)) ; mse # [1] 0.07925

## F value for the treatment, so blocking is a factor.
F0 <- msf/mse ; F0 # [1] 75.89485

## To be symmetric, the F value for the blocking
msb/mse # [1] 21.11356

## Verify with functions.
m <- aov(y~a+b,d) ; summary(m)
v <- anova(m)
sum(v['Sum Sq'])

## Modle Diagnostics
plot(m)

```