# Introduction to Engineering Statistics and R 
Engineering Statistics (IMSE 4410) Spring 2016. 
Copyright 2013-2016 by Timothy Middelkoop License CC by SA 3.0

## Overview
* The fundamental concepts in the class are statistical models and variance.
* Regression and ANOVA are two ways of looking at variance in data.  
* From this we can ask the question: "Can I explain this data using a model?"
* This class will give you an understanding of the fundamentals, it is not a modeling class.
 * (there are too many ways to model, to test, data)
* There is a whole degree for this class, we will have to gloss over a lot of material
* Understand the core.  Ask Questions. Do not guess.

## Workflow
* Attend the lectures.
* Understand at home with the notes.
* Do the homework (ungraded w/ solutions).
* Ask questions.

### Why R?
* Spreadsheets suck.
* Built for statistics.
* Industrial grade, finance, bioinformatics use it.
* Works similar to Matlab
* Free and open source.
* Large community, lots of community support, resources, extensions.
* Smart people use it
* Good connectivity with databases and spreadsheets.
* Fairly modern.
* Automated workflow.
* R Studio.
* Did I say Spreadsheets suck.

### Lets get started.
* R will be used as a big calculator in this class.
* The project will extensively use R.
* Install R and RStudio (see ReadMe.md in https://github.com/MiddelkoopT/Stats-2016-Spring)
* Lecture notes are a running commentary the R calcuations (R session or RMarkdown)

```{r}
# The basics.
3+4
## [1] 7

# The '#' symbol is used for comments
# the '##' symbol also a comment and is the result of a command   
# [1] is a line number
# 7 is the sum of 3+4

# numbers are reals, not integers, just like a calculator
7/3

# we even have a memory button
3+4 -> m
m

# some of us are used to seeing it in the other direction
# doing more than one thing with the ;
m <- 3+4 ; m

# we have function buttons
sin(1)

# and more interesting ones. 

# Let's use this as an example of classroom work (board work)
# Problem: How to add a sequence of numbers
# Example: add 11, 12, 13, and 14.
# Solution: 11+12+13+14=50

# Add a sequence of numbers (and the hand caclulated expected result)
11+12+13+14
# 50

# this could get old fast.

# stats is about lots of data, so lets store some in an array
a <- c(11,12,13,14) ; a

# R has one based arrays (R is for humans) [] is an index/position
a[1]

a[5]
# oops, nothing there (Not Available)

# NA is not zero, its Not Available
NA+1

# back to the task at hand, compute the solution
sum(a)

# More data please, but I don't like to type
# from http://www.cyclismo.org/tutorial/R/input.html
d <- read.csv("http://www.cyclismo.org/tutorial/R/_static/simple.csv",header=TRUE)
d

# top is the column names (from headers=TRUE)
# left is rows, just the row number/index.

# just like an array
d[1,2]

# I forgot the name of the column
names(d)

# This is stats class so lets take the mean
mean(d$velocity)

# the $ symbol access the column

# quick check our answers
12+14+8+10+13+11
68/6

# now we have an answer, lets not type in numbers unless we have to (a fundamenal rule)
sum(d$velocity)

# we do not divide it by 6... no more typing remember.
sum(d$velocity)/length(d$velocity)

# Yes, we know how mean and sum work now.  
# Always do things the long way first, to make sure it is doing what you expect.
# This is not a black box.  We have a help button
# ?mean

# we can do other things with tables
names(d)
# [1] "trial"    "mass"     "velocity"

# I wonder if there is a relationship between mass and velocity
d$velocity/d$mass

# vectors are cool if you did not guess the operation is element wise.
# always verify
d$velocity[1]/d$mass[1]

# I understand what this is, do you? Verify!
d$velocity
d$mass
d$velocity[1] 
d$mass[1]
12/10

# yes, looks good to me.

# Interesting but how about more visual [plot(x,y)]
plot(d$mass,d$velocity)

# There is a nicer way of looking at this [y~x]
plot(velocity~mass,d)
```


### Extra/Graduate
Often I include extra advanced R/Statistics.  Graduate students are responsible for this material.

```{r}
# we can create a table (data.frame) ourselves.
d <- data.frame(x=c(11,12,13),y=c(21,22,23))
d

# note our example is not symmetric.
d[2,1]

# yes that is what we expected.

# rows
d[1,]

# columns
d[,2]

# which is named y
d$y

# can also name rows
rownames(d) <- c("A","B","C")
d

# we can use this for indexing as well
d['C','y']

# Yes naming is as expected
colnames(d)

# And that is it.
```

### Summary of Session
* numbers
* operators
* memory and assignment
* arrays
* functions
* data.table

## Workflow

Before class:

* Read the chapter.
* Lecture notes limit scope and support the caclcuations.

During Class:

* Present the problem, theory, and background
* Develop a simple example.
* Replicate calculations using R
* Show the R way of doing it.

After Class:

* In class notes will be the in class R session capture.
* Do the homework
* Review posted solution only after you have worked the problems.

## References
* http://cran.r-project.org/doc/manuals/R-intro.html
* http://www.cyclismo.org/tutorial/R/

