# Engineering Statistics
IMSE 4410 Spring 2016. Copyright 2013-2016 by Timothy Middelkoop License CC by SA 3.0

## Hypothesis Testing - Chapter 9

"You can't prove anything, you can only disprove it" H0

### Type I Error

* Type I error:  Reject H0 when H0 is true.
* P(Type I error):  $\alpha$ (alpha) error or significance level.

```{r}
# Example 9-1.2
n <- 10
sigma <- 2.5
mu <- 50
sigmaxb <- sigma/sqrt(n) ; sigmaxb ## [1] 0.7905694
```

$\sigma_{\bar x}$ (sigmaxb) is the standard deviation of the sample mean $\bar x$ (xb)

```{r}
# find the alpha error
alpha <- pnorm(48.5,mean=mu,sd=sigmaxb,lower.tail=TRUE) +
         pnorm(51.5,mean=mu,sd=sigmaxb,lower.tail=FALSE) ; alpha ## [1] 0.05777957
```

This means alpha (5.7%) of all random samples would lead to the rejection of the null hypothesis H0:mu==50.

### Type II Error

* Type II error: Fail to reject H0 when H0 is false.
* P(Type II error): $\beta$ (beta) error
* 1-$\beta$: power of a test.
* When we fail to reject H0 it does not mean H0 is true (it is just not, not true)

Note, since H0 is false we need to know what H0 actually is. We say "under H1" what it actually is.

**under H0: $\mu=50$, under H1: $\mu=52$ (which says $\mu$ not 50 but 52)**

```{r}
pnorm(51.5,mean=52,sd=sigmaxb,lower.tail=TRUE) ## [1] 0.2635446
pnorm(48.5,mean=52,sd=sigmaxb,lower.tail=TRUE) ## [1] 4.77346e-06

beta <- pnorm(51.5,mean=52,sd=sigmaxb,lower.tail=TRUE) - 
  	pnorm(48.5,mean=52,sd=sigmaxb,lower.tail=TRUE) ; beta ## [1] 0.2635399

# Do not like the probability? Increase n 
n <- 16
sigmaxb <- sigma/sqrt(n) ; sigmaxb

# Remember sigmaxb is the standard deviation on the sample mean, not the population!
pnorm(51.5,mean=52,sd=sigmaxb,lower.tail=TRUE) ## [1] 0.2118554 # better

alpha <- pnorm(48.5,mean=mu,sd=sigmaxb,lower.tail=TRUE) +
		pnorm(51.5,mean=mu,sd=sigmaxb,lower.tail=FALSE) ; alpha ## [1] 0.01639507

# but look what happened to our alpha value.
```

### The relationship between Type I and Type II

Type I and Type II errors are related, reduce one, increase the other.
The size of the critical region determines alpha.
An increase in $n$, with $\alpha$ (alpha) constant, decreases $\beta$ (beta).
When H0 is false and the true value approaches H0, beta increases.

* We control alpha.
* Rejecting H0 is a strong conclusion.
* We typically want to reject H0 leaving us with H1 being true.

* One sided tests, the direction matters.
* What do we want "strong evidence" of?  Rejecting H0 will indicate strong evidence.
* For the bursting strength example, H0=200, H1>200 (Eq 9-5) indicates strong evidence that the bottles meet the specifications.

### p-values

Don't report just rejection ($p<\alpha$), report how strong is the rejection (the p-value)

**How much alpha would it take (the p-value) to reject H0.**

```{r}
# 9-1.4
# H0: mu=50; H1: mu!=50
mu <- 50 ; n <- 16 ; sigma <- 2.5
xb <- 51.3
sigmaxb <- sigma/sqrt(n) ; sigmaxb ## [1] 0.625

# find the symmetric limit (sxb)
sxb <- mu - (xb-mu) ; sxb ## [1] 48.7

# Add the tails to find the p-value (p)
p <- pnorm(xb,mean=mu,sd=sigmaxb,lower.tail=FALSE) +
   pnorm(sxb,mean=mu,sd=sigmaxb,lower.tail=TRUE) ; p ## [1] 0.03752553
```

So p-value < 0.05 so we reject H0, the p-value is the observed significance level.
Any p-value > 0.0375 would lead to rejecting H0.  _Think about it._

The hypothesis test is either True or False, we are just stating the significance level.

**Review: "Acceptance Region" table on page 291 (9-1.2)**

```{r}
# 48.5 < xb < 51.5, n=10
sigma <- 2.5; mu0 <- 50 

# alpha
n <- 10
sigmaxb <- sigma/sqrt(n)
pnorm(48.5,mean=mu0,sd=sigmaxb,lower.tail=TRUE) +
pnorm(51.5,mean=mu0,sd=sigmaxb,lower.tail=FALSE)   ## [1] 0.05777957

mu1 <- 52
pnorm(51.5,mean=mu1,sd=sigmaxb,lower.tail=TRUE) -
pnorm(48.5,mean=mu1,sd=sigmaxb,lower.tail=TRUE)   ## [1] 0.2635399

mu1 <- 50.5
pnorm(51.5,mean=mu1,sd=sigmaxb,lower.tail=TRUE) -
pnorm(48.5,mean=mu1,sd=sigmaxb,lower.tail=TRUE)   ## [1] 0.8913424

# Rest are easy, lets try n=16, H1:m=52
mu1 <- 52
n <- 16
sigmaxb <- sigma/sqrt(n)
pnorm(51.58,mean=mu1,sd=sigmaxb,lower.tail=TRUE) -
pnorm(48.42,mean=mu1,sd=sigmaxb,lower.tail=TRUE)   ## [1] 0.2507918
```


### References

* "Computing the Standard Deviation of Sample Means," https://www.utdallas.edu/~metin/Ba3352/qch9-10.pdf

