# Introduction to Engineering Statistics and R 
Engineering Statistics (IMSE 4410) Spring 2016. 
Copyright 2013-2015 by Timothy Middelkoop License CC by SA 3.0

## Descriptive Statistics - Chapter 6

### Sample Mean (6-1)
Simple descriptions of data.
```{r}
x <- c(12.6,12.9,13.4,12.3,13.6,13.5,12.6,13.1)
# Verify the size of the data.
length(x)
# View the data to get a feel of it.
plot(x)
# Verify the analisys.
sum(x)
sum(x)/length(x)
# Do it the R way.
mean(x)
```
This the sample mean, which is different from the population mean!

What is the population mean?  Can we estimate this from the sample mean? It turns out Yes.

$$\mu=\sum_{i=1}^N x_i$$

mu=sum(xi)/N

### Sample Variance (6-1)
What's this?

Sample Variance is defined as.  Note the Sample and the s^2.

$$s^2 = {\sum_{i=1}^n (x_i-\bar x)^2 \over n-1}$$

s2=sum((xi-xb)^2)/(n-1) 

Calculation:
```{r}
x
xb <- mean(x) ; xb

# compute and verify
x-xb
# Test your knoledge
sum(x-xb)
# sqared, note the extra brackets.
sum((x-xb)^2)
# this is why...
sum(x-xb)^2
# OH.

# Keep this around
n <- length(x) ; n

# skip a step
s2 <- sum((x-xb)^2)/(n-1) ; s2
## 0.2285714

## note the similarity between our equation above and what we just did.

# We will not be using the "short" hand calculation way in this class.

## sample standard deviation.
s <- sqrt(s2) ; s
## 0.4780914

# Verify with text.

# Verfy manual
# ?sd

# The R Way
sd(x)

```

### Population Variance (6-1)

Population Variance and Sample variance are NOT this same.

Population Variance:
$$\sigma^2 = {\sum_{i=1}^N (x_i-\bar x)^2 \over N}$$

Note the differences, including n.

### Degrees of Freedom. (6-1)

The term (n-1) is used in the sample since we use $\bar x$ (xb).  This is also because if we use this we would consistently underestimate $s^2$.  This is also because of the concept of _degrees of freedom_, we can determine any one term if we remove it (calculate backwards).

### Sample Range.

For completeness the sample range r=max(x)-min(x), which is practically R code and is 1.3.
